### 2021-04-08 ch04_papers_ecs_tcr_assessment/papers_ecs_tcr_assessment.tex
Spell- and grammer-checked chapter 4

To reduce uncertainties in multi-model climate projections, a first important
step is the assessment of the desired target variables in the corresponding
climate model ensemble. This is particularly relevant when a new generation of
\acp{ESM} is published that includes considerable modifications compared to the
respective predecessor model versions. Currently, we are in a situation like
this with the \acs{CMIP}6 ensemble, for which new data is still released every
day (as of \TheMonth{}). In light of the upcoming international climate
assessment of the \ac{IPCC}'s \acs{AR}6, the evaluation of policy-relevant
metrics like \ac{ECS} or \ac{TCR} for these new \acs{CMIP}6 models is crucial
since they provide vital information about the future climate of planet Earth.
As the \ac{ESMValTool} is an excellent tool that allows for a quick and robust
evaluation of new \ac{CMIP} data as soon as it gets published on the \ac{ESGF}
servers (see \cref{sec:03:esmvaltool}), it is used here to assess and analyze
\ac{ECS} and \ac{TCR} of the \acs{CMIP}6 models and compare it to corresponding
results of predecessor model generations and international climate assessments.
This work is already published in two scientific papers: \textcite{Bock2020}
and \textcite{Meehl2020}. For \textcite{Bock2020}, the author of this thesis
contributed their figure 8 (a bar chart showing the \ac{ECS} for several
\acs{CMIP}3, \acs{CMIP}5, and \acs{CMIP}6 models), their figure 10 (map plots
showing cloud-related feedback parameters for the \acs{CMIP}5 and \acs{CMIP}6
\acp{MMM}; see \cref{fig:04:cloud_feedback_parameters}), code to create these
figures with the \ac{ESMValTool}, and text for the manuscript (in particular a
section about the evaluation of \ac{ECS}). For \textcite{Meehl2020}, the author
of this thesis contributed all figures and tables, code to produce these with
the \ac{ESMValTool}, and text for the manuscript. This
\namecref{ch:04:papers_ecs_tcr_assessment} first presents the evaluation of
climate sensitivity for the \acs{CMIP}5 and \acs{CMIP}6 ensembles
(\cref{sec:04:evaluation_ecs_and_tcr}) and puts the corresponding \ac{ECS} and
\ac{TCR} values into historical context (\cref{sec:04:historical_context}).
Finally, possible reasons for the apparent increase of climate sensitivity in
the \acs{CMIP}6 models are discussed in detail
(\cref{sec:04:possible_reasons_high_ecs_cmip6}).

@@

Following \cref{subsec:02:ecs}, \ac{ECS} is calculated with the Gregory
regression method using \ac{GSAT} and \ac{TOA} net radiation data for 150 years
of a \nxcotwo{4} simulation \autocite{Gregory2004}. This calculation is
illustrated in \cref{fig:04:gregory_regression_cmip6_mmm} for the \acs{CMIP}6
\ac{MMM}, which yields \iac{ECS} of $3.74 \unit{K}$ when all 150 years of the
run are used. However, similar to the \acs{CMIP}5 ensemble, the exact value of
\ac{ECS} depends on the years considered in the Gregory regression (see
\cref{fig:02:gregory_regression_different_years}). Using only the first 20
years of the simulation gives a significantly lower \ac{ECS} of $3.31 \unit{K}$
than using only the last 130 years of the simulation, which gives \iac{ECS} of
$4.05 \unit{K}$. As thoroughly describes in \cref{subsec:02:ecs}, the reason
for this is the state and time dependence of the climate feedback parameter,
which is given by the slope of the Gregory regression line. Due to non-linear
effects in the feedbacks, this slope changes over time, resulting in lower
(higher) values of \ac{ECS} when only early (late) years of the simulation are
considered. In contrast to that, the transient response of the climate system
\ac{TCR} is calculated from \ac{GSAT} data at the time of \ac{CO2} doubling in
a \onepctcotwo{} run \commentcite{Bindoff2013}{see \cref{subsec:02:tcr}}. Using
these two definitions and the \ac{ESMValTool}, \ac{ECS} and \ac{TCR} are
assessed for all \acs{CMIP}5 and \acs{CMIP}6 models where the necessary
temperature and radiation data is available (as of \TheMonth{}).
\Cref{tab:04:ecs_tcr_cmip5,tab:04:ecs_tcr_cmip6} give an overview over the
results for the \acs{CMIP}5 and \acs{CMIP}6 models, respectively.

@@

\begin{table}[!t]
  \centering
  \csvreader[ECSAndTCRTable]{
    ch04_papers_ecs_tcr_assessment/data/cmip5_ecs-tcr.csv}{}{
    \dataset & $\idx$ & $\ecs$ & $\tcr$}
  \caption[
    \Acf{ECS} and \acf{TCR} calculated for the \acs{CMIP}5 models.
  ]{
    \Acf{ECS} and \acf{TCR} calculated for the \acs{CMIP}5 models. Details on
    this are given in \cref{subsec:02:ecs} and \cref{subsec:02:tcr},
    respectively. The \acf{MMM} is calculated from the Gregory regression
    method using the \acs{MMM} net \acf{TOA} radiation and the \acs{MMM} change
    in \acf{GSAT} similar to \cref{fig:04:gregory_regression_cmip6_mmm}. The
    multi-model standard deviation is given by the sample standard deviation of
    \acs{ECS} evaluated over all climate models (using the normalization $1 /
    M$, where $M$ is the number of models). Corresponding references for each
    model are given in \cref{tab:app:a:cmip5_models}. \AdaptedFrom{Meehl2020}.
  }
  \label{tab:04:ecs_tcr_cmip5}
\end{table}

@@

The first striking feature of these two tables is the increased \ac{MMM} of
\ac{ECS} and \ac{TCR} in \acs{CMIP}6. For \ac{ECS}, the \acs{CMIP}6 \ac{MMM} is
about $16 \unit{\%}$ ($0.51 \unit{K}$) higher than the corresponding
\acs{CMIP}5 \ac{MMM}. For \ac{TCR}, the relative difference between the two
model ensembles is notably smaller with about $10 \unit{\%}$ ($0.19 \unit{K}$).
The spread in the multi-model ensembles (expressed as the multi-model standard
deviation) shows an even larger increase in the \acs{CMIP}6 ensemble: For
\ac{ECS}, the relative difference is about $49 \unit{\%}$ ($0.72 \unit{K}$ in
\acs{CMIP}5 to $1.07 \unit{K}$ in \acs{CMIP}6), and for \ac{TCR}, the relative
difference is about $17 \unit{\%}$ ($0.36 \unit{K}$ in \acs{CMIP}5 to $0.42
\unit{K}$ in \acs{CMIP}6). The main reason for the increased \ac{MMM} and
spread in the \acs{CMIP}6 ensemble is the existence of several models with very
high values of \ac{ECS} and \ac{TCR}. In addition to that, there also exist a
number of models with very low values of \ac{ECS}. Therefore, the \acs{CMIP}6
model ranges of \ac{ECS} and \ac{TCR} are well outside the corresponding
assessed ranges given by the latest published \ac{AR} of the \ac{IPCC} from
2013 \autocite{Stocker2013} with $\rangeunit{1.8}{5.6}{K}$ for \ac{ECS}
(\acs{AR}5: $\rangeunit{1.5}{4.5}{K}$) and $\rangeunit{1.3}{3.0}{K}$
(\acs{AR}5: $\rangeunit{1.0}{2.5}{K}$) for \ac{TCR}.

@@

To illustrate the results of the previous
\namecref{sec:04:evaluation_ecs_and_tcr} and put them into historical context,
\cref{fig:04:historical_ecs_tcr_ranges} shows the assessed \ac{ECS} and
\ac{TCR} ranges over the years from the Charney report \autocite{Charney1979}
and the different \acp{AR} of the \ac{IPCC} in combination with the
corresponding modeled ranges from the different \ac{CMIP} generations. Since
the Charney report in 1979, the assessed range of \ac{ECS} of
$\rangeunit{1.5}{4.5}{K}$ has almost remained unchanged during the last 40
years \autocite{Charney1979, Mitchell1990, Kattenberg1996, Albritton2001,
  Stocker2013} with the exception of \acs{AR}4, where the lower bound was
temporarily increased to $2.0 \unit{K}$ \autocite{Solomon2007}. For \acs{AR}2,
\acs{AR}4, and \acs{AR}5, the corresponding climate model generations from
\acs{CMIP}1, \acs{CMIP}3, and \acs{CMIP}5 more or less agree with this assessed
range \autocite{Kattenberg1996, Randall2007, Flato2013}. In contrast to that,
some of the early climate models used in \acs{AR}1 and some \acs{CMIP}2 models
used in \acs{AR}3 exhibit values of \acs{ECS} well above $4.5 \unit{K}$,
resulting in an upper model range of about $0.6 \unit{K}$ higher than the
assessed range \autocite{Mitchell1990, Cubasch2001}. However, these deviations
from the assessed range are small compared to the ones present in the
\acs{CMIP}6 ensemble, which have become evident over the course of 2020. As
shown in \cref{fig:04:historical_ecs_tcr_ranges}, both the lower and upper
bound are more extreme than in any previous climate model generation, resulting
in a \acs{CMIP}6 \ac{ECS} model range of $\rangeunit{1.8}{5.6}{K}$ (as of
\TheMonth{}). This increase in range is particularly relevant for the upper
bound, which is about $1.1 \unit{K}$ higher than the assessed upper bound of
$4.5 \unit{K}$. Moreover, it is not only a small fraction of models that exceed
this upper bound: out of the 42 analyzed \acs{CMIP}6 models, a third (14
models) have \iac{ECS} above $4.5 \unit{K}$. In addition to the analysis
presented here, several modeling centers independently confirmed high \ac{ECS}
values in their models \autocite{Andrews2019, BodasSalcedo2019, Gettelman2019,
  Wyser2020}. A list of possible reasons for the increased climate sensitivity
in the \acs{CMIP}6 ensemble including aerosol-cloud interactions and changes in
the shortwave cloud feedback over the Southern Ocean is given in
\cref{sec:04:possible_reasons_high_ecs_cmip6}.

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch04_papers_ecs_tcr_assessment/figs/historical_ecs_tcr_ranges.pdf}
  \caption[
    Historical values of the \acf{ECS} and the \acf{TCR} for different
    \aclp{AR} and climate model ensembles.
  ]{
    Assessed values of the \acf{ECS} (blue bars) and the \acf{TCR} (red bars)
    over the years in the Charney report from 1979 \autocite{Charney1979} and
    the subsequent \acfp{AR} of the \acf{IPCC} \autocite{Mitchell1990,
      Kattenberg1996, Albritton2001, Solomon2007, Stocker2013}. Orange and
    green bars represent the modeled ranges of \acs{ECS} and \acs{TCR},
    respectively, from the different climate model ensembles of the \acf{CMIP}
    with the corresponding \acf{MMM} illustrated with the horizontal black
    lines (the green circle represents output from a single climate model). The
    numbers correspond to individual \acs{CMIP}5 and \acs{CMIP}6 models (see
    \cref{tab:04:ecs_tcr_cmip5,tab:04:ecs_tcr_cmip6}). \AdaptedFrom{Meehl2020}.
  }
  \label{fig:04:historical_ecs_tcr_ranges}
\end{figure}

In contrast to \ac{ECS}, \ac{TCR} has only been evaluated since \acs{AR}1 in
1990, in which it is estimated with $2.3 \unit{K}$ using only a single climate
model \commentcite{Bretherton1990}{see single green circle in
  \cref{fig:04:historical_ecs_tcr_ranges}}. Over the years, the modeled range
of \ac{TCR} has decreased from about $\rangeunit{1.3}{3.8}{K}$ in the
\acs{CMIP}1 models \autocite{Kattenberg1996} to about $\rangeunit{1.1}{2.6}{K}$
in the \acs{CMIP}5 ensemble \autocite{Flato2013}. Similarly, the assessed range
of \ac{TCR}, which was first quoted in \acs{AR}4 with $\rangeunit{1.0}{3.0}{K}$
\autocite{Solomon2007}, has been reduced to $\rangeunit{1.0}{2.5}{K}$ in the
subsequent \acs{AR}5 \autocite{Stocker2013}. While the corresponding climate
model ranges from \acs{CMIP}3 and \acs{CMIP}5 agree with these assessed ranges
\autocite{Randall2007, Flato2013}, the upper bound of the \acs{CMIP}6 model
range ($\rangeunit{1.3}{3.0}{K}$) is well above the assessed upper bound of
$2.5 \unit{K}$ given by \acs{AR}5. However, unlike \ac{ECS}, also the lower
bound of the \acs{CMIP}6 model range increased compared to the \acs{CMIP}5
ensemble, leading to a similar \ac{TCR} model spread in both climate model
generations. Due to the correlation of \ac{TCR} and \ac{ECS} given by
\cref{eq:02:tcr_vs_ecs}, the increase of \ac{TCR} in many \acs{CMIP}6 models is
not surprising in light of the many high \ac{ECS} models that are present the
\acs{CMIP}6 ensemble.

@@

Another possible reason for the high climate sensitivity in the \acs{CMIP}6
ensemble is related to cloud feedbacks \autocite{Bock2020}. As already
discussed in \cref{subsec:02:climate_feedbacks}, uncertainties in cloud
feedbacks are a major source of uncertainty of \ac{ECS} in modern \acp{ESM}
\autocite{Boucher2013, Flato2013}. Thus, changes in processes that are related
to cloud feedbacks immediately impact the climate models' climate
sensitivities. The particular feedback mechanism that is thought to be relevant
for the high \ac{ECS} in \acs{CMIP}6 is connected to cloud phase changes over
the Southern Ocean that are present in earlier \ac{CMIP} generations (\eg{}
\acs{CMIP}5). This so-called \emph{cloud phase change feedback} is illustrated
in \cref{fig:04:cloud_phase_change_feedback} (green arrows). As the climate
warms, the predominantly ice clouds over the Southern Ocean in these climate
models become liquid clouds \autocite{McCoy2015}. As a result, the cloud will
get more reflective to the incoming solar radiation since a cloud consisting of
smaller liquid droplets reflects more sunlight than a cloud consisting of
larger ice crystals (for a fixed water content), which has a cooling effect on
the climate system. In addition, predominantly liquid clouds tend to
precipitate less than mixed-phase clouds (composed of liquid water and ice).
This leads to a higher liquid water content of the cloud and a further
amplification of the cooling effect. Overall, this forms a strong negative
shortwave cloud feedback over the Southern Ocean in climate models from
previous \ac{CMIP} generations.

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch04_papers_ecs_tcr_assessment/figs/cloud_phase_change_feedback.pdf}
  \caption[
    Schematic illustration of the strong negative shortwave cloud phase change
    feedback over the Southern Ocean.
  ]{
    Schematic illustration of the strong negative shortwave cloud phase change
    feedback over the Southern Ocean, which is present in earlier \acs{CMIP}
    model generations like \acs{CMIP}5 (green arrows). Due to improvements in
    the microphysical representation of mixed-phase clouds, some \acs{CMIP}6
    models simulate a higher fraction of (supercooled) liquid water in the
    clouds over the present-day Southern Ocean than their corresponding
    predecessor models \autocite{BodasSalcedo2019, Gettelman2019}. Thus, in the
    affected \acs{CMIP}6 models, the strong negative cloud feedback that
    results from a cloud phase change from ice clouds in the present-day to
    liquid clouds in the future is substantially reduced (orange arrows). This
    leads to an increase in the global net cloud feedback
    \autocite{BodasSalcedo2019, Tan2016} and the \acf{ECS}
    \autocite{Andrews2019, Gettelman2019}, which is a possible explanation for
    the high \acs{ECS} values found in some \acs{CMIP}6 models.
  }
  \label{fig:04:cloud_phase_change_feedback}
\end{figure}

However, it is well known that the low-level mixed-phase clouds over the
Southern Ocean in these earlier climate models are biased towards too large
amounts of ice crystals and too little amounts of supercooled liquid water when
compared to satellite observations \autocite{BodasSalcedo2016}. Thus, the size
of the resulting strong negative cloud phase change feedback in these models
has long been questioned \autocite{McCoy2015, Tan2016}. Improvements in the
representation of mixed-phase clouds are known to reduce this long-standing
cloud phase bias over the Southern Ocean \autocite{BodasSalcedo2016, McCoy2016}
and to improve the representation of both cloud microphysical structure and
cloud radiative impacts in this region \autocite{Hyder2018, Kay2016}. For this
reason, the microphysical representation of mixed-phase clouds has been updated
in some \acs{CMIP}6 models. For example, in \acs{CESM}2, a new mixed-phase ice
nucleation scheme by \textcite{Hoose2010} replaces the \textcite{Meyers1992}
empirical scheme used in predecessor models \autocite{Gettelman2019}.
Simulations with the \textcite{Meyers1992} scheme show almost no supercooled
liquid water in high-latitude clouds and a strong negative cloud phase change
feedback when ice turns to liquid. On the contrary, simulations with the new
ice nucleation scheme by \textcite{Hoose2010} show a strong reduction in this
feedback. Since the new scheme also reduces significant cloud phase biases over
the Southern Ocean that are present in the old scheme, the reduced negative
cloud phase change feedback is thought to be more consistent with observations
of the current climate \autocite{Gettelman2019}. Another example is the
\acs{HadGEM}3 model, for which a new cloud scheme that accounts for turbulent
production of liquid water in mixed-phase clouds has been implemented
\autocite{BodasSalcedo2019}. The new scheme parameterizes the role of
subgrid-scale turbulence in the production and maintenance of supercooled
liquid water \autocite{Furtado2016}. As a consequence, the cloud liquid water
path over the Southern Ocean is increased, which is more in line with
observations from microwave satellite retrievals. Moreover, the increase in the
cloud liquid water path with warming is reduced in \acs{HadGEM}3, leading to a
reduction in the negative cloud phase change feedback over the Southern Ocean
\autocite{BodasSalcedo2019}. In addition, \textcite{Williams2020} show that
short-term forecasts created with \acs{HadGEM}3 are more accurate than
comparable simulations from predecessor model versions (using traditional
metrics from numerical weather prediction), which indicates an improvement of
the physical processes in the new climate model.

>>>>

To reduce uncertainties in multi-model climate projections, a first important
step is the assessment of the desired target variables in the corresponding
climate model ensemble. This is particularly relevant when a new generation of
\acp{ESM} is published that includes considerable modifications compared to the
respective predecessor model versions. Currently, we are in a situation like
this with the \acs{CMIP}6 ensemble, for which new data is still released every
day (as of \TheMonth{}). In light of the upcoming international climate
assessment of the \ac{IPCC}'s \acs{AR}6, the evaluation of policy-relevant
metrics like \ac{ECS} or \ac{TCR} for these new \acs{CMIP}6 models is crucial
since they provide vital information about the future climate of the planet
Earth. As the \ac{ESMValTool} is an excellent tool that allows for a quick and
robust evaluation of new \ac{CMIP} data as soon as it gets published on the
\ac{ESGF} servers (see \cref{sec:03:esmvaltool}), it is used here to assess and
analyze \ac{ECS} and \ac{TCR} of the \acs{CMIP}6 models and compare it to
corresponding results of predecessor model generations and international
climate assessments. This work is already published in two scientific papers:
\textcite{Bock2020} and \textcite{Meehl2020}. For \textcite{Bock2020}, the
author of this thesis contributed their figure 8 (a bar chart showing the
\ac{ECS} for several \acs{CMIP}3, \acs{CMIP}5, and \acs{CMIP}6 models), their
figure 10 (map plots showing cloud-related feedback parameters for the
\acs{CMIP}5 and \acs{CMIP}6 \acp{MMM}; see
\cref{fig:04:cloud_feedback_parameters}), code to create these figures with the
\ac{ESMValTool}, and text for the manuscript (in particular a section about the
evaluation of \ac{ECS}). For \textcite{Meehl2020}, the author of this thesis
contributed all figures and tables, code to produce these with the
\ac{ESMValTool}, and text for the manuscript. This
\namecref{ch:04:papers_ecs_tcr_assessment} first presents the evaluation of
climate sensitivity for the \acs{CMIP}5 and \acs{CMIP}6 ensembles
(\cref{sec:04:evaluation_ecs_and_tcr}) and puts the corresponding \ac{ECS} and
\ac{TCR} values into historical context (\cref{sec:04:historical_context}).
Finally, possible reasons for the apparent increase of climate sensitivity in
the \acs{CMIP}6 models are discussed in detail
(\cref{sec:04:possible_reasons_high_ecs_cmip6}).

@@

Following \cref{subsec:02:ecs}, \ac{ECS} is calculated with the Gregory
regression method using \ac{GSAT} and \ac{TOA} net radiation data for 150 years
of a \nxcotwo{4} simulation \autocite{Gregory2004}. This calculation is
illustrated in \cref{fig:04:gregory_regression_cmip6_mmm} for the \acs{CMIP}6
\ac{MMM}, which yields \iac{ECS} of $3.74 \unit{K}$ when all 150 years of the
run are used. However, similar to the \acs{CMIP}5 ensemble, the exact value of
\ac{ECS} depends on the years considered in the Gregory regression (see
\cref{fig:02:gregory_regression_different_years}). Using only the first 20
years of the simulation gives a significantly lower \ac{ECS} of $3.31 \unit{K}$
than using only the last 130 years of the simulation, which gives \iac{ECS} of
$4.05 \unit{K}$. As thoroughly describes in \cref{subsec:02:ecs}, the reason
for this is the state and time dependence of the climate feedback parameter,
which is given by the slope of the Gregory regression line. Due to non-linear
effects in the feedbacks, this slope changes over time, resulting in lower
(higher) values of \ac{ECS} when only early (late) years of the simulation are
considered. In contrast to that, the transient response of the climate system
\ac{TCR} is calculated from \ac{GSAT} data at the time of \ac{CO2} doubling in
a \onepctcotwo{} run \commentcite{Bindoff2013}{see \cref{subsec:02:tcr}}. Using
these two definitions and the \ac{ESMValTool}, \ac{ECS} and \ac{TCR} are
assessed for all \acs{CMIP}5 and \acs{CMIP}6 models where the necessary
temperature and radiation data are available (as of \TheMonth{}).
\Cref{tab:04:ecs_tcr_cmip5,tab:04:ecs_tcr_cmip6} give an overview of the
results for the \acs{CMIP}5 and \acs{CMIP}6 models, respectively.

@@

\begin{table}[!t]
  \centering
  \csvreader[ECSAndTCRTable]{
    ch04_papers_ecs_tcr_assessment/data/cmip5_ecs-tcr.csv}{}{
    \dataset & $\idx$ & $\ecs$ & $\tcr$}
  \caption[
    \Acf{ECS} and \acf{TCR} calculated for the \acs{CMIP}5 models.
  ]{
    \Acf{ECS} and \acf{TCR} calculated for the \acs{CMIP}5 models. Details on
    this are given in \cref{subsec:02:ecs} and \cref{subsec:02:tcr},
    respectively. The \acf{MMM} is calculated from the Gregory regression
    method using the \acs{MMM} net \acf{TOA} radiation and the \acs{MMM} change
    in \acf{GSAT} similar to \cref{fig:04:gregory_regression_cmip6_mmm}. The
    multi-model standard deviation is given by the sample standard deviation of
    \acs{ECS} calculated over all climate models (using the normalization $1 /
    M$, where $M$ is the number of models). Corresponding references for each
    model are given in \cref{tab:app:a:cmip5_models}. \AdaptedFrom{Meehl2020}.
  }
  \label{tab:04:ecs_tcr_cmip5}
\end{table}

@@

The first striking feature of these two tables is the increased \ac{MMM} of
\ac{ECS} and \ac{TCR} in \acs{CMIP}6. For \ac{ECS}, the \acs{CMIP}6 \ac{MMM} is
about $16 \unit{\%}$ ($0.51 \unit{K}$) higher than the corresponding
\acs{CMIP}5 \ac{MMM}. For \ac{TCR}, the relative difference between the two
model ensembles is notably smaller with about $10 \unit{\%}$ ($0.19 \unit{K}$).
The spread in the multi-model ensembles (expressed as the multi-model standard
deviation) shows an even larger increase in the \acs{CMIP}6 ensemble: For
\ac{ECS}, the relative difference is about $49 \unit{\%}$ ($0.72 \unit{K}$ in
\acs{CMIP}5 to $1.07 \unit{K}$ in \acs{CMIP}6), and for \ac{TCR}, the relative
difference is about $17 \unit{\%}$ ($0.36 \unit{K}$ in \acs{CMIP}5 to $0.42
\unit{K}$ in \acs{CMIP}6). The main reason for the increased \ac{MMM} and
spread in the \acs{CMIP}6 ensemble is the existence of several models with very
high values of \ac{ECS} and \ac{TCR}. In addition to that, there also exist
some models with very low values of \ac{ECS}. Therefore, the \acs{CMIP}6 model
ranges of \ac{ECS} and \ac{TCR} are well outside the corresponding assessed
ranges given by the latest published \ac{AR} of the \ac{IPCC} from 2013
\autocite{Stocker2013} with $\rangeunit{1.8}{5.6}{K}$ for \ac{ECS} (\acs{AR}5:
$\rangeunit{1.5}{4.5}{K}$) and $\rangeunit{1.3}{3.0}{K}$ (\acs{AR}5:
$\rangeunit{1.0}{2.5}{K}$) for \ac{TCR}.

@@

To illustrate the results of the previous
\namecref{sec:04:evaluation_ecs_and_tcr} and put them into historical context,
\cref{fig:04:historical_ecs_tcr_ranges} shows the assessed \ac{ECS} and
\ac{TCR} ranges over the years from the Charney report \autocite{Charney1979}
and the different \acp{AR} of the \ac{IPCC} in combination with the
corresponding modeled ranges from the different \ac{CMIP} generations. Since
the Charney report in 1979, the assessed range of \ac{ECS} of
$\rangeunit{1.5}{4.5}{K}$ has almost remained unchanged during the last 40
years \autocite{Charney1979, Mitchell1990, Kattenberg1996, Albritton2001,
  Stocker2013}, with the exception of \acs{AR}4, where the lower bound was
temporarily increased to $2.0 \unit{K}$ \autocite{Solomon2007}. For \acs{AR}2,
\acs{AR}4, and \acs{AR}5, the corresponding climate model generations from
\acs{CMIP}1, \acs{CMIP}3, and \acs{CMIP}5 more or less agree with this assessed
range \autocite{Kattenberg1996, Randall2007, Flato2013}. In contrast to that,
some of the early climate models used in \acs{AR}1 and some \acs{CMIP}2 models
used in \acs{AR}3 exhibit values of \acs{ECS} well above $4.5 \unit{K}$,
resulting in an upper model range of about $0.6 \unit{K}$ higher than the
assessed range \autocite{Mitchell1990, Cubasch2001}. However, these deviations
from the assessed range are small compared to the ones present in the
\acs{CMIP}6 ensemble, which have become evident over the course of 2020. As
shown in \cref{fig:04:historical_ecs_tcr_ranges}, both the lower and upper
bound are more extreme than in any previous climate model generation, resulting
in a \acs{CMIP}6 \ac{ECS} model range of $\rangeunit{1.8}{5.6}{K}$ (as of
\TheMonth{}). This increase in range is particularly relevant for the upper
bound, which is about $1.1 \unit{K}$ higher than the assessed upper bound of
$4.5 \unit{K}$. Moreover, it is not only a small fraction of models that exceed
this upper bound: out of the 42 analyzed \acs{CMIP}6 models, a third (14
models) have \iac{ECS} above $4.5 \unit{K}$. In addition to the analysis
presented here, several modeling centers independently confirmed high \ac{ECS}
values in their models \autocite{Andrews2019, BodasSalcedo2019, Gettelman2019,
  Wyser2020}. A list of possible reasons for the increased climate sensitivity
in the \acs{CMIP}6 ensemble including aerosol-cloud interactions and changes in
the shortwave cloud feedback over the Southern Ocean is given in
\cref{sec:04:possible_reasons_high_ecs_cmip6}.

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch04_papers_ecs_tcr_assessment/figs/historical_ecs_tcr_ranges.pdf}
  \caption[
    Historical values of the \acf{ECS} and the \acf{TCR} for different
    \aclp{AR} and climate model ensembles.
  ]{
    Assessed values of the \acf{ECS} (blue bars) and the \acf{TCR} (red bars)
    over the years in the Charney report from 1979 \autocite{Charney1979} and
    the subsequent \acfp{AR} of the \acf{IPCC} \autocite{Mitchell1990,
      Kattenberg1996, Albritton2001, Solomon2007, Stocker2013}. Orange and
    green bars represent the modeled ranges of \acs{ECS} and \acs{TCR},
    respectively, from the different climate model ensembles of the \acf{CMIP}
    with the corresponding \acf{MMM} illustrated with the horizontal black
    lines (the green circle represents the output from a single climate
    model). The numbers correspond to individual \acs{CMIP}5 and \acs{CMIP}6
    models (see \cref{tab:04:ecs_tcr_cmip5,tab:04:ecs_tcr_cmip6}).
    \AdaptedFrom{Meehl2020}.
  }
  \label{fig:04:historical_ecs_tcr_ranges}
\end{figure}

In contrast to \ac{ECS}, \ac{TCR} has only been evaluated since \acs{AR}1 in
1990, in which it is estimated with $2.3 \unit{K}$ using only a single climate
model \commentcite{Bretherton1990}{see single green circle in
  \cref{fig:04:historical_ecs_tcr_ranges}}. Over the years, the modeled range
of \ac{TCR} has decreased from about $\rangeunit{1.3}{3.8}{K}$ in the
\acs{CMIP}1 models \autocite{Kattenberg1996} to about $\rangeunit{1.1}{2.6}{K}$
in the \acs{CMIP}5 ensemble \autocite{Flato2013}. Similarly, the assessed range
of \ac{TCR}, which was first quoted in \acs{AR}4 with $\rangeunit{1.0}{3.0}{K}$
\autocite{Solomon2007}, has been reduced to $\rangeunit{1.0}{2.5}{K}$ in the
subsequent \acs{AR}5 \autocite{Stocker2013}. While the corresponding climate
model ranges from \acs{CMIP}3 and \acs{CMIP}5 agree with these assessed ranges
\autocite{Randall2007, Flato2013}, the upper bound of the \acs{CMIP}6 model
range ($\rangeunit{1.3}{3.0}{K}$) is well above the assessed upper bound of
$2.5 \unit{K}$ given by \acs{AR}5. However, unlike \ac{ECS}, also the lower
bound of the \acs{CMIP}6 model range increased compared to the \acs{CMIP}5
ensemble, leading to a similar \ac{TCR} model spread in both climate model
generations. Due to the correlation of \ac{TCR} and \ac{ECS} given by
\cref{eq:02:tcr_vs_ecs}, the increase of \ac{TCR} in many \acs{CMIP}6 models is
not surprising in light of the many high \ac{ECS} models that are present in
the \acs{CMIP}6 ensemble.

@@

Another possible reason for the high climate sensitivity in the \acs{CMIP}6
ensemble is related to cloud feedbacks \autocite{Bock2020}. As already
discussed in \cref{subsec:02:climate_feedbacks}, uncertainties in cloud
feedbacks are a major source of uncertainty of \ac{ECS} in modern \acp{ESM}
\autocite{Boucher2013, Flato2013}. Thus, changes in processes that are related
to cloud feedbacks immediately impact the climate models' climate
sensitivities. The particular feedback mechanism that is thought to be relevant
for the high \ac{ECS} in \acs{CMIP}6 is connected to cloud phase changes over
the Southern Ocean that are present in earlier \ac{CMIP} generations (\eg{}
\acs{CMIP}5). This so-called \emph{cloud phase change feedback} is illustrated
in \cref{fig:04:cloud_phase_change_feedback} (green arrows). As the climate
warms, the predominantly ice clouds over the Southern Ocean in these climate
models become liquid clouds \autocite{McCoy2015}. As a result, the cloud will
get more reflective to the incoming solar radiation since a cloud consisting of
smaller liquid droplets reflects more sunlight than a cloud consisting of
larger ice crystals (for fixed water content), which has a cooling effect on
the climate system. In addition, predominantly liquid clouds tend to
precipitate less than mixed-phase clouds (composed of liquid water and ice).
This leads to a higher liquid water content of the cloud and a further
amplification of the cooling effect. Overall, this forms a strong negative
shortwave cloud feedback over the Southern Ocean in climate models from
previous \ac{CMIP} generations.

\begin{figure}[t]
  \centering
  \includegraphics[width=\LargeFigureWidth{}]{
    ch04_papers_ecs_tcr_assessment/figs/cloud_phase_change_feedback.pdf}
  \caption[
    Schematic illustration of the strong negative shortwave cloud phase change
    feedback over the Southern Ocean.
  ]{
    Schematic illustration of the strong negative shortwave cloud phase change
    feedback over the Southern Ocean, which is present in earlier \acs{CMIP}
    model generations like \acs{CMIP}5 (green arrows). Due to improvements in
    the microphysical representation of mixed-phase clouds, some \acs{CMIP}6
    models simulate a higher fraction of (supercooled) liquid water in the
    clouds over the present-day Southern Ocean than their corresponding
    predecessor models \autocite{BodasSalcedo2019, Gettelman2019}. Thus, in the
    affected \acs{CMIP}6 models, the strong negative cloud feedback that
    results from a cloud phase change from ice clouds in the present day to
    liquid clouds in the future is substantially reduced (orange arrows). This
    leads to an increase in the global net cloud feedback
    \autocite{BodasSalcedo2019, Tan2016} and the \acf{ECS}
    \autocite{Andrews2019, Gettelman2019}, which is a possible explanation for
    the high \acs{ECS} values found in some \acs{CMIP}6 models.
  }
  \label{fig:04:cloud_phase_change_feedback}
\end{figure}

However, it is well known that the low-level mixed-phase clouds over the
Southern Ocean in these earlier climate models are biased towards too large
amounts of ice crystals and too little amounts of supercooled liquid water when
compared to satellite observations \autocite{BodasSalcedo2016}. Thus, the size
of the resulting strong negative cloud phase change feedback in these models
has long been questioned \autocite{McCoy2015, Tan2016}. Improvements in the
representation of mixed-phase clouds are known to reduce this long-standing
cloud phase bias over the Southern Ocean \autocite{BodasSalcedo2016, McCoy2016}
and to improve the representation of both cloud microphysical structure and
cloud radiative impacts in this region \autocite{Hyder2018, Kay2016}. For this
reason, the microphysical representation of mixed-phase clouds has been updated
in some \acs{CMIP}6 models. For example, in \acs{CESM}2, a new mixed-phase ice
nucleation scheme by \textcite{Hoose2010} replaces the \textcite{Meyers1992}
empirical scheme used in predecessor models \autocite{Gettelman2019}.
Simulations with the \textcite{Meyers1992} scheme show almost no supercooled
liquid water in high-latitude clouds and a strong negative cloud phase change
feedback when ice turns to liquid. On the contrary, simulations with the new
ice nucleation scheme by \textcite{Hoose2010} show a strong reduction in this
feedback. Since the new scheme also reduces significant cloud phase biases over
the Southern Ocean that are present in the old scheme, the reduced negative
cloud phase change feedback is thought to be more consistent with observations
of the current climate \autocite{Gettelman2019}. Another example is the
\acs{HadGEM}3 model, for which a new cloud scheme that accounts for the
turbulent production of liquid water in mixed-phase clouds has been implemented
\autocite{BodasSalcedo2019}. The new scheme parameterizes the role of
subgrid-scale turbulence in the production and maintenance of supercooled
liquid water \autocite{Furtado2016}. As a consequence, the cloud liquid water
path over the Southern Ocean is increased, which is more in line with
observations from microwave satellite retrievals. Moreover, the increase in the
cloud liquid water path with warming is reduced in \acs{HadGEM}3, leading to a
reduction in the negative cloud phase change feedback over the Southern Ocean
\autocite{BodasSalcedo2019}. In addition, \textcite{Williams2020} show that
short-term forecasts created with \acs{HadGEM}3 are more accurate than
comparable simulations from predecessor model versions (using traditional
metrics from numerical weather prediction), which indicates an improvement of
the physical processes in the new climate model.
